---
title: "Probability - Tasks"
author: "Florian Oswald"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Task 1: Probabilities

Given a fair 6-sided dice 🎲, i.e. an outcome space $\Omega = \{1,2,3,4,5,6\}$
and events

$$\begin{align}
A &= \{2,4,6\},\\
B &= \{1,3,5\}.
\end{align}$$

Define $X$ as the outcome of rolling the dice once. What is $\Pr(X \in A)$?

## Answer

In this simple sample space (i.e. a sample space where each element has the same probability of occuring), we can just count. There are 6 elements in $\Omega$, event $A$ contains 3 of those, i.e. in 3 out of 6 cases $X$ could land on an even number. That is

$$\Pr(X \in A) = \frac{\text{number of success}}{\text{all potential outcomes}} = \frac{3}{6} = 0.5$$


---

class: inverse

# Task 2: 2 Dice

Given two fair 6-sided dice 🎲 🎲, what is the probability of obtaining *at least* once the face "5"?

Again, counting. The total number of events from 2 dice are

```{r}
o = outer(1:6,1:6,FUN = function(x,y){paste0("(",x,",",y,")")})
o
```
i.e. for each potential value of the first dice (`1,2,...,6`) we have 6 values for the second dice (`1,2,...,6`). In total, 36 potential outcomes. 

How many of those correspond to *at least* once the face "5"? Well, just look for "5" above! (Caution, don't count "(5,5)" twice!)

So, we have 

```{r}
o[,5]
```

and 

```{r}
o[5,]
```
which together would make 12 elements, but we must not count "(5,5)" twice. Therefore,

$$\Pr(\text{At least once "5"}) = \frac{11}{36}$$

# Task: Computing Variance by hand

1. Compute the variance of one 6-sided dice!

$$Var(X) = E(X^2) - \mu_x^2$$

we know that $\mu_x = 3.5 = \frac{7}{2}$ from before, so just need the first part.

$$E(X^2) = \sum_{i=1}^6 \Pr(X = i) i^2 = \frac{1}{6} 91$$
Therefore $$Var(X) = E(X^2) - \mu_x^2 =  \frac{91}{6} - \frac{49}{4} = 11.6667$$


2. compute the variance of $X \sim \text{Bernoulli}(p)$!

Again, 

$$Var(X) = E(X^2) - \mu_x^2 = E(X^2) - p^2$$
$E(X^2) = p 1^2 + (1-p) 0^2 = p$. this gives us

$$Var(X) = E(X^2) - \mu_x^2 = p - p^2 = p(1-p)$$
Bonus question: what value of $p$ maximizes this variance?

```{r,echo = FALSE}
x = seq(0,1,length.out = 100)
y = x * (1-x)

plot(x,y,type = "l")
```


